What should I notice? Evaluating the memorability of events for abductive inference.

### 2021-10-04 paper.tex
Adding modifications from JL's version

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{url}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage[ruled,linesnumbered]{algorithm2e}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}
\title{
  What should I notice? Evaluating memorability of events for abductive inference.
}
\author{
  \IEEEauthorblockN{
  \IEEEauthorrefmark{1}\IEEEauthorrefmark{2} Étienne Houzé,
  \IEEEauthorrefmark{1} Jean-Louis Dessalles,
  \IEEEauthorrefmark{1} Ada Diaconescu,
  \IEEEauthorrefmark{2} David Menga
  }

@@

\begin{abstract}
  When confronted to an unprecedented situation, humans typically show good
  performance in quickly identifying noticeable past events and proposing them
  as possible causal hypotheses. This kind of abductive inference is widely
  overlooked in modern AI approaches which rely on massive datasets to learn
  associative patterns. Our proposal is to formalize and compute a ``memorability''
  score over a memory of various recorded events from a cyber-physical system.
 This score can later be used either to select only relevant information to be
 remembered, or to propose causal hypotheses in unusual situations, on demand.
 As such, the approach aims at being complementary to more traditional
 learning-focused techniques. We provide theoretical ground for our approach by
 using and extending existing results and ideas from Algorithmic Information
 Theory and provide an implementation example, showing practical results in a
 smart-home scenario.
\end{abstract}

\begin{IEEEkeywords}
  Complexity, Algorithmic Information Theory, Simplicity, Abduction, Surprise
\end{IEEEkeywords}

@@

As a user has just turned on the TV in her all-equipped living-room, the
lights dim and the window blinds lower. Intrigued by this behavior, she
quickly infers that both light dimming and the blinds closing occurred as a
consequence of the TV being turned on. How did she come to this conclusion?
By performing \emph{abductive inference} \cite{magnani_abduction_2011}, which
is a key element of humans' ability to understand the world: from the
observed consequences, infer the possible causes.

In this example, there are mainly three possibilities to come to the
conclusion. If the user knows how the smart living-room system works, if she
knows the set of applied rules or parameters, she has access to a causal
knowledge she can use for the abduction. Otherwise, if she has no knowledge
about the system but the behavior occurs frequently, examining past
correlations will reveal that the television frequently causes the blinds to
close and the lights to dim. Else, if this is the first instance of the TV
being turned on in this living-room, there is no previous occurrence upon
which the user may rely to draw conclusions. However, she remains able to
infer the TV as a possible cause for the observed reaction: by noticing the
TV as a memorable recent event (since it is its first occurrence), she
proposes it as a possible cause for the questioned phenomenon. This example
shows how humans are able to use distinct methods to perform abductive tasks
and infer new knowledge. While abduction of the first two kinds can be
automated using knowledge bases and statistical approaches, this paper
proposes an approach to evaluate the memorability of events, allowing
machines to perform ``memorability-based'' abduction.

@@

The difficulty of implementing this kind of abduction for machines comes from
several factors. First, events can be of different nature, and not directly
comparable. In fact, even for specific systems such as smart homes, events
range from device removal to a presence detection or an unusually high
temperature. How can one can then assess which ones are more memorable?
Furthermore, even for comparable events, different characteristics can be put
forward to argue for the most memorable: is a record-high temperature 47 days
ago more memorable than the small deviation recorded just 3 minutes ago? To
this day, no current system proposes to use all this information from various
event types from different devices to compute a unified metrics of
``memorability''.

@@

To tackle this challenge, we rely on the following observation: while all
events, regardless of their characteristics or nature, can be uniquely
described using a combination of qualifiers, the most memorable ones are
likely to require less words to be described. Think, for instance, of
``last year's hottest day'' and ``the 182nd day from 7 years ago'': the
former seems simpler to describe. To quantify the notion, we can evaluate the complexity of each
description, taking into account both the complexity of concept words (a date
of occurrence, a temperature ranking), and the arguments (the hottest, the
182nd, 7). The resulting value defines the  \emph{description complexity}
of the events. As our intuition is that memorable events require simpler and less numerous qualifiers to be
unambiguously described than other unremarkable ones, they will stand out as simpler in terms of description complexity.

@@

\begin{figure}[ht]
  \centering
\includegraphics[width=\linewidth]{figures/filters}
\caption{Retrieving an event through successive predicative
filters. From the base memory (yellow), successive filters select events satisfying the associated predicate (grey arrows). For example,
   filter $f_1$ selects events from last year, i.e. which satisfy the
   predicate $\mathtt{date}(event, 1\_year)$. In this case, successively
   applying filters $f_1$, $f_2$ and $f_3$ yields a unique event $e_h$, last
   year's hottest day. The complexity of this event can then be upper bounded
   by the complexity of the three filters, since they give a unambiguous way
to describe the event in the memory.}
  \label{fig:filters}
\end{figure}

For machines to implement and compute description complexity, we need a formal canvas and computation methods which coincide with human intuition . To do so, we rely on Algorithmic Information Theory, which also appears to be consistent with the human perception of complexity\cite{li_introduction_2008,dessalles2011coincidences,delahaye_numerical_2012}. Our method is as follows: we
consider events as being unordered elements from a base \emph{memory}. To
reproduce the language features applicable to events, we use \emph
{predicates}, i.e. functions assigning a boolean value to an event and
eventual arguments. For instance, the predicate $\mathtt{date}
(\cdot, 1\_year)$ assigns \emph{true} to events that occurred last year. Selecting
all events, from the memory, that satisfy a given predicate corresponds to
a \emph{filter} operation, and yields another memory, a subset of the former.
This filtering operation can then be repeated, each iteration, selecting
fewer events, until a singleton memory is reached, meaning that the sequence
of predicates can \emph{retrieve} unambiguously the unique remaining event.
The description complexity of this event can thus be upper-bounded by the
number of bits required to describe the filters used in the retrieval
process. Figure \ref{fig:filters} illustrates this evaluation for the event
$e_h$: ``last year's hottest day''.

@@

The rest of this paper is organized as follows: we first briefly introduce
some notions of Algorithmic Information Theory and its applications in
section~\ref{sec:theory}, then we explicit our methods to compute the
memorability of past events in cyber-physical systems~\ref
{sec:computing}. We illustrate this approach by providing an implementation
relying on a few predicates on events from a smart home simulation~
(section \ref{sec:example}). In section~\ref{sec:related} we briefly review
other related works using complexity theory or trying to explain smart homes
and how our work can be linked with them before exploring potential
extensions of our work in section \ref{sec:future}

\section{Theoretical Background}
\label{sec:theory}
Kolmogorov complexity formally quantifies the amount of information required
for the computation of a finite binary string\footnote{While the definition still
holds for some infinite binary strings (think of the representation of the
decimals of $\pi$), we restrict ourselves to finite strings in this paper.} (or
any object represented by a finite binary
string)\cite{kolmogorov_three_1965,li_introduction_2008}. For such a binary
string $s$, its complexity $K(s)$, is the bit length of the shortest program $p$
which, if given as input to a universal Turing Machine $U$, outputs $s$.
\begin{equation}
  \label{eq:kolmo_def}
  K_{U}(s) = \min_{p}\left\{l(p)|U(p)=s\right\}
\end{equation}

The first notable property of this definition is its universality: while the
choice of the Turing machine $U$ used for the computations appears in the
definition of eq. \ref{eq:kolmo_def}, all results stand, up to an additional
constant. Think, for instance, how any Turing-complete programming language can
be turned into any other language, using a compiler program. In fact, since any
Turing machine $U'$ can be emulated by $U$ from a
finite program $p_{U}$, we have the following inequality:
\begin{equation}
  \label{eq:inequality_univ}
  K_{U}(s) \le l(p_{u}) + K_{U}(s)
\end{equation}

From this first result, we can then define a universal complexity $K(s)$, which
no longer depends on the choice of a Turing machine $U$, such that, for any
machine $U$,
\begin{equation}
  \forall U\in\text{TM}, \forall s, |K(s) - K_{u}(s)| \le C_{U}
\end{equation}
where the additional constant $C_{U}$ does not depend on the object $s$.

It is important to note that the notion of Kolmogorov complexity is not related
to computation complexity, as there is no requirement on the execution time of
the programs, only their length in bits matters for the computation of
complexity. In fact, it can be shown that Kolmogorov complexity is not
computable\cite{li_introduction_2008}. The sketch of the proof is the
following: since the definition of complexity from Equation. \ref
{eq:kolmo_def} requires to find the shortest of all programs on a given Turing
machine outputting $s$, it requires to run all programs up to a given length
and compare their result to $s$. However, being able to do so means that we
would be able to tell whether these programs terminate. Since the termination
of programs is famously undecidable, by extension, the computation of
Kolmogorov complexity is impossible. One can however easily approximate it with
upper bounds, by exhibiting a program outputting $s$.

Interestingly, Kolmogorov's definition of complexity matches the intuitive
notion and perception of complexity from a human standpoint. For instance, the
complexity of short binary strings evaluated in \cite{delahaye_numerical_2012}
shows similar results to human perception of complex strings and patterns. More
recently, \cite{murena_solving_2020} used Kolmogorov complexity to solve
analogies and showed results close to human expectations.

This bridge between Algorithmic Information Theory and human perception of
complexity is extended to define the notion of simplicity and unexpectedness,
which are considered of uttermost importance in cognitive science\cite
{chater_simplicity_2003}.
\cite{dessalles2011coincidences} proposes a formal definition of the
 unexpectedness $U(e)$ of an event, as the difference between an a-priori
 expected causal complexity $C_{w}(e)$ and the actual observed complexity $C
 (e)$.
\begin{equation}
\label{eq:unexpected} U(e) = C_{w}(e) - C(e)
\end{equation}
This result comes from the understanding that, while Kolmogorov complexity is
computed with Turing machine, it can be used as an elegant proxy for modeling
information processing in the human brain, and thus helps design a notion of
simplicity or complexity of events.

The definition of Equation \ref{eq:unexpected} allows to model phenomenons
such as coincidences: imagine that you happen to run into someone in a park.
If this person has no particular link to you, the event will be quite
trivial: the complexity of describing this person will be equivalent to
distinguishing it from the global population, which will be roughly
equivalent as the complexity of describing that this person happens to be in
the same park as you. On the other hand, if you run into your best friend in
a park, as the complexity of describing your best friend is significantly
lower, the description complexity $C(e)$ drops while the causal complexity
remains $C_W (e)$ unchanged. Therefore the event becomes unexpected.

Using these insights from AIT, we define the memorability $M(e)$ of an event as the absolute difference between the description complexity $K_d(e)$ of an event and its expected description complexity $K_{exp}(e)$:
\begin{equation}
  \label{eq:memorability}
  M(e) = |K_{exp}(e) - K_d(e)|
\end{equation}
Contrary to the definition of unexpectedness from Equation \ref{eq:unexpected}, we use an absolute value: this is done to acknowledge events more complex than expected as memorable\footnote{In the original paper \cite{dessalles2011coincidences}, unusually complex events could be handled by considering complexity itself as a way to describe the event: see ``the Pisa Tower effect''\cite{dessalles_pisa_nodate}}. In the next section, we formally define the description complexity $K_d$ and the expected complexity $K_{exp}$ of events and detail how we can compute approximations of these quantities.

@@

Computing the memorability of events begins with the introduction of formal definitions of events, how they are stored, how we can describe them and how these descriptions can be used to retrieve them.

We define \emph{events} as data points augmented with a \emph{label} indicating their nature (temperature event, failure event, addition/removal of a device) and a timestamp of occurrence. Formally:
\begin{equation}
  \label{eq:event}
  e = (l, t,\mathcal{D})
\end{equation}
where $l$ is the label, $t$ the timestamp and $\mathcal{D}$ a multi-dimensional data point representing the various characteristics of $e$: its duration, the maximum temperature reached, the sensor name, its position, etc. Labels can also be considered as classes of events, of which each event is a particular instance.

To model how humans are able to describe events by using qualifiers, we use \emph{predicates}: function operating on events, eventually taking additional parameters, returning a boolean value: $\pi(e, a_1, a_2, \dots, a_n) \mapsto \{O,1\}$ is a predicate of arity $n$ operating on event $e$. In the rest of this paper, we will prefer the equivalent notation $\pi(e, k) \mapsto \{0,1\}$, where $k$ is a binary string encoding the sequence of arguments $a_1, \dots, a_n$. Using this notation, the predicate $\pi$ becomes a boolean function operating on $\mathbf{E} \times \{0,1\}^*$:
\begin{equation}
  \label{eq:predicate}
  \pi : \begin{cases}
    \mathcal{M}\times \{0,1\}^{*} &\mapsto \{0,1\} \\
    (e, k) &\mapsto \pi_{k}(e)
    \end{cases}
\end{equation}
An example of predicate is to take $\pi = \mathtt{year}$ and $k$ a string encoding the number $1$, thus
constructing the predicate $\mathtt{year(}e, 1\mathtt{)}$, which tells whether
the event $e$ occurred $1$ years ago.

As events occur, they are stored in a \emph{memory} $M_0$. As they are not directly comparable, the memory $M_0$ can be considered as having the structure of an unordered set. We denote by $\mathcal{M}$ the subsets of $M_0$. By extension, elements of $\mathcal{M}$, i.e. subsets of $M_0$, are also called \emph{memories}.

By applying a given predicate $\pi_k$ to all events contained in a memory $M \subseteq M_0$, and selecting only events satisfying $\pi_k$ yields another memory $M_1 \subseteq M \subseteq M_0$. We call this operation a \emph{filter}:
\begin{equation}
  \label{eq:filter}
f_{\pi, k}: \begin{cases}
  \mathcal{M} & \mapsto \mathcal{M}             \\
  M           & \mapsto \{e \in M | \pi_k(e) \}
\end{cases}
\end{equation}
For instance, using the same $\pi = \mathtt{year}$ and $k=1$ as above, we can
build the filter $f_{\pi, k} = \mathtt{last\_{}year}$, which selects all events
that occurred last year.

As the output of a filter applied to a memory $M$ is another memory
object $M' \subseteq M$, we can compose filter
functions. A sequence of such filters is called a \emph{retrieval path}
\begin{equation}
\label{eq:ret_def}
p = (f_{\pi_{1}, k_{1}}, \dots, f_{\pi_{n}, k_{n}})
\end{equation}
and by definition
$p(M) = f_{\pi_{n}, k_{n}}(\dots(f_{\pi_{1}, k_{1}}(M)))$.
In case the result of the operation $p(M)$ contains a single element
$e$, we say that the path $p$ \emph{retrieves} the element $e$ from $M$, and write
$p(M) = e$. In the example shown in Figure \ref{fig:filters}, the three filters $f_1, f_2, f_3$ form a retrieval path retrieving the event ``last year's hottest day'' from the base memory $M_0$.

@@

By considering only a finite number of possible predicates $\pi$ and arguments $k$, and a maximum path length, we can construct a finite set $P$ of possible retrieval paths. By limiting the search over this set, we can upper bound the description complexity, and use this upper bound as an approximation:

@@

The approximation of description complexity from Equation \ref{eq:approx_k_desc} allows for
a direct implementation, which is shown in Algorithm \ref{alg:complex_iter}. This
algorithm operates iteratively: starting with the base memory $M_0$
(line 1), we apply all possible predicate concepts $\pi$ from a given finite set
$\Pi$ and programs $k$ (lines 6-7), up to a given length $\mathtt
{max\_len}$ bits, and apply them: $M' = f_{\pi, k}(M)$ (line 12). We then store
the pairs $(M', \mathtt{len(}\pi, k\mathtt{)})$ in an array $\mathtt{future_
{explore}}$. At the end of the iteration, the results of the filters become the
memories which will be explored during the next iteration(lines 21--23). Each
pass thus explore retrieval paths of increasing length. When a singleton memory
is reached, the complexity of its unique element is upper-bounded with the
length of the corresponding retrieval path (line 14).

\subsection{Computing Memorability}
As stated in Equation \ref{eq:memorability}, we define memorability $M(e)$ as the absolute difference between the description complexity of an event and its expected value. As we've just defined $K_d(e)$ and provided an approximation in Equation \ref{eq:approx_k_desc}, we now focus on defining the description expected complexity of an event, $K_{exp}(e)$.

As intended in the Equation \ref{eq:memorability}, this term evaluates the complexity the user, or the system, would expect the event $e$ to have without actually describing it, based on their previous knowledge. In our canvas, this prior knowledge consists of the base memory $M_0$. The expected complexity of the event $e$ can be computed with a simple first-order approximation, i.e. estimating the average complexity of ``similar events'' over the base memory $M_0$.

Still, difficulty remains in the definition of what
can be considered \emph{similar} events. Given that we deal with non comparable events, the only possibility to universally define the notion of similarity is to once again refer to \emph{predicates}. Thus, for a given event $e$, and a given predicate $\pi_k$, we define a $\pi_k$ neighborhood of $e$ as the set $N_{\pi, k}(e)$ of all other events satisfying $\pi_k$.

@@

This definition is consistent with the intuitive ideas that more similar events should weight more in the computation. Indeed, if $e'$
is \emph{very} similar to $e$, it will appear in many neighborhoods, since it
satisfies mostly the same predicates as $e$. Therefore, it will
be present in more terms in Equation \ref{eq:expected}, and will weight more in
the final result.

\subsection{From memorability to abduction}
% On peut étendre la définition utilisée pour la surprise en faisant intervenir
% les prédicats'' `sympathiques` par rapport à ce que l'on questionne.
The problem of the abductive inference is different from the basic computation
of the memorability score. Here, \emph{knowing} that we want to find a cause $c$
to an observed effect $e$, we try to find the most remarkable event in past
memory in that regard. While our ``memorability'' score aims to identify the
most remarkable events, it does not take into account the added knowledge about
the effect $e$ that is available for abductive inference.

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{url}
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage[ruled,linesnumbered]{algorithm2e}
\usepackage{xcolor}
\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}
\title{
What should I notice? Evaluating the memorability of events for abductive inference.
}
\author{
  \IEEEauthorblockN{
\IEEEauthorrefmark{1}\IEEEauthorrefmark{2} Étienne Houzé,
\IEEEauthorrefmark{1} Jean-Louis Dessalles,
\IEEEauthorrefmark{1} Ada Diaconescu,
\IEEEauthorrefmark{2} David Menga
  }

@@

\begin{abstract}
  When confronted to an unprecedented situation, humans typically show good
  performance in quickly identifying noticeable past events and proposing them
  as possible causal hypotheses. This kind of abductive inference is widely
  overlooked in modern AI approaches which rely on massive datasets to learn
  associative patterns. Our proposal is to formalize and compute a ``memorability''
score over a set of events recorded from a cyber-physical system.
This score can then be used to select relevant information to be
remembered and to propose causal hypotheses in unusual situations on demand.
The approach is meant to be complementary to more traditional
learning-focused abduction techniques. We provide theoretical ground for our approach
based on Algorithmic Information
Theory. We also provide an implementation example, a
smart-home scenario, to show how our approach could translate in practice.
\end{abstract}

\begin{IEEEkeywords}
Kolmogorov complexity, Algorithmic Information Theory, Simplicity, Abduction, Surprise
\end{IEEEkeywords}

@@

As a user has just turned on the TV in her all-equipped living-room, the
lights dim and the window blinds go down. Intrigued by this behavior, she
quickly infers that both light dimming and blind closing occurred as a
consequence of the TV being turned on. How did she come to this conclusion?
By performing \emph{abductive inference} \cite{magnani_abduction_2011}. This mental operation
is a key element of humans' ability to understand the world: from the
observed consequences, they infer the possible causes.

In this example, there are mainly three possibilities to come to the
conclusion. (1) If the user knows how the smart living-room system works, if she
knows the underlying rules or parameters, she may use this causal
knowledge to perform abduction. (2) If she has no knowledge
about the system but made several observation of the same behavior, she may examine past
correlations and figure out that turning on the TV set often leads the blinds to
close and the lights to dim. (3) If there are no previous occurrences of the event (e.g. it is the first time she turns on the TV
in the living-room), she may still be able to
suspect that the TV is a possible cause for the observed event, just because it appears to her as a memorable recent event (as it is its first occurrence). This example
suggests that human beings are able to use distinct methods to perform abductive tasks
and infer new knowledge. While the first two mechanisms can be
automated using knowledge bases and statistical methods, ``memorability-based'' abduction needs to be investigated in more depth.

@@

Defining a score of memorability is not straightforward. First, events can be of different nature, and not directly
comparable. Even for restricted systems such as smart homes, noticeable events
range from device removal to presence detection or unusually high
temperatures. Even for comparable events, the problem is to weigh different characteristics: is a record-high temperature 47 days
ago more memorable than the small deviation recorded just 3 minutes ago? To
our knowledge, no current system proposes to combine various
event types from different devices to compute a unified metrics of
``memorability''.

@@

To address this issue, we started from the following intuition: while all
events, regardless of their characteristics or nature, can be uniquely
described using a combination of quantitative or qualitative qualifiers, the most memorable ones are
likely to require less words to be described. ``Last year's hottest day'' is simpler than ``the 182\textsuperscript{nd} day 7 years ago''. How can we quantify this relative simplicity? We could evaluate the complexity of each
description, taking into account both the complexity of the concept words (a date
of occurrence, a temperature ranking), and of the arguments (1st hottest,
182 and 7). This approach, The resulting values define the  \emph{description complexity}
of events. Our intuition is that memorable events require simpler and less numerous qualifiers to be
unambiguously described than unremarkable ones.

@@

\begin{figure}[ht]
  \centering
\includegraphics[width=\linewidth]{figures/filters}
\caption{Retrieving an event through successive predicative
filters. From the base memory (yellow), successive filters select events satisfying the associated predicate (grey arrows). For example,
filter $f_1$ selects events from last year, i.e. which satisfy the
predicate $\mathtt{date}(event, 1\_year)$. In this case, successively
applying filters $f_1$, $f_2$ and $f_3$ yields a unique event $e_h$, last
year's hottest day. The complexity of this event can then be upper-bounded
by the complexity of the three filters, since they give an unambiguous way
to describe the event within the memory.}
  \label{fig:filters}
\end{figure}

For machines to implement and compute description complexity, we need a formal framework and computation methods that are in line with human intuition.  Algorithmic Information Theory (AIT) appears to be such a framework, as it is consistent with the human perception of complexity\cite{li_introduction_2008,dessalles2011coincidences,delahaye_numerical_2012}. Our method is as follows: we
consider events as being elements stored in what we call \emph{base memory}. To
reproduce the language features applicable to events, we use \emph
{predicates}, i.e. functions assigning a boolean value to events. For instance, the predicate $\mathtt{date}
  (\cdot, 1\_year)$ is \emph{true} of events that occurred last year. Selecting
all events, from the memory, that satisfy a given predicate corresponds to
a \emph{filter} operation. It generates another memory that is a subset of the previous one.
The filtering operation can then be repeated, selecting
fewer events at each iteration, until a singleton memory is reached. This means that the sequence
of predicates could unambiguously \emph{retrieve} the unique remaining event.
The description complexity of this event can thus be upper-bounded by the
number of bits required to describe the filters used in the retrieval
process. Figure \ref{fig:filters} illustrates this process for the event: ``last year's hottest day''.

@@

We will first briefly introduce
some basic notions of Algorithmic Information Theory. Then, in
section~\ref{sec:computing}, we show how we compute the
memorability of past events. We illustrate our approach in~
(section \ref{sec:example}) by describing an implementation
in a smart home simulation context. In section~\ref{sec:related} we review
a few studies that use complexity theory to generate explanations for the smart home
and we discuss how our work can be linked with them. We eventually explore potential
extensions of our work in section \ref{sec:future}

\section{Theoretical Background}
\label{sec:theory}
Kolmogorov complexity formally quantifies the amount of information required
for the computation of a finite binary string\footnote{Though the definition
holds for some infinite binary strings (think of the representation of the
decimals of $\pi$), we restrict ourselves here to finite strings.} (or
any object represented by a finite binary
string)\cite{kolmogorov_three_1965,li_introduction_2008}. The complexity $K(s)$ of a (finite) binary string is the length in bits of the shortest program $p$
which, if given as input to a universal Turing Machine $U$, outputs $s$.
\begin{equation}
  \label{eq:kolmo_def}
  K_{U}(s) = \min_{p}\left\{l(p)|U(p)=s\right\}
\end{equation}

The first notable property of this definition is its universality: while the
choice of the Turing machine $U$ used for the computations appears in the
definition of eq. \ref{eq:kolmo_def}, all results hold, up to an additional
constant, if we change the machine. Think how any Turing-complete programming language can
be turned into any other language, using an interpreter or a compiler program. Since any
Turing machine $U'$ can be emulated by $U$ from a
finite program $p_{U}$, we have the following inequality:
\begin{equation}
  \label{eq:inequality_univ}
K_{U'}(s) \le l(p_{u}) + K_{U}(s)
\end{equation}

From this first result, we can then define complexity $K(s)$, based on the choice of a reference Turing machine, such that, for any other
machine $U$ taken from the set $\text{TM}$ of Turing machines:
\begin{equation}
  \forall U\in\text{TM}, \forall s, |K(s) - K_{u}(s)| \le C_{U}
\end{equation}
where the additional constant $C_{U}$ does not depend on $s$.

Note that the notion of Kolmogorov complexity involves no requirement on the execution time of
programs, only their length in bits matters for the computation of
complexity. Though Kolmogorov complexity can be shown to be
incomputable\cite{li_introduction_2008},
%The sketch of the proof is the
%following: since the definition of complexity from Equation. \ref
%{eq:kolmo_def} requires to find the shortest of all programs on a given Turing
%machine outputting $s$, it requires to run all programs up to a given length
%and compare their result to $s$. However, being able to do so means that we
%would be able to tell whether these programs terminate. Since the termination
%of programs is famously undecidable, by extension, the computation of
%Kolmogorov complexity is impossible.
it can be approximated with
upper bounds by exhibiting a program outputting $s$.

Interestingly, Kolmogorov complexity matches the intuitive
notion and perception of complexity from a human standpoint. For instance, the
complexity of short binary strings evaluated in \cite{delahaye_numerical_2012}
shows similar results to human perception of complex strings and patterns. More
recently, \cite{murena_solving_2020} used Kolmogorov complexity to solve
analogies and showed results close to human expectations.

The bridge between Algorithmic Information Theory (AIT) and human perception of
complexity can be pushed farther thanks to the notions of simplicity and unexpectedness,
which are sometimes regarded of uttermost importance in cognitive science\cite
{chater_simplicity_2003}.
\cite{dessalles2011coincidences} proposes a formal definition of the
unexpectedness $U(e)$ of an event, as the difference between an a-priori
expected causal complexity $K_{w}(e)$ and the actual observed complexity $K
  (e)$.
\begin{equation}
\label{eq:unexpected} U(e) = K_{w}(e) - K(e)
\end{equation}

This result comes from the understanding that, while Kolmogorov complexity is
ideally computed using a Turing machine, it can be used as a proxy for modeling
information processing in the human brain, and thus can be used to define a notion of
simplicity or complexity of events.

Definition \ref{eq:unexpected} allows to model phenomena
such as coincidences: imagine that you happen to run into someone in a park.
If this person has no particular link to you, the event will be quite
trivial: the complexity of describing this person will be equivalent to
distinguishing it from the global population, which is also roughly
equivalent to the (causal) complexity of describing the circumstances having brought this person to be in
that park as the same time as you. On the other hand, if you run into your best friend in
a park, as the complexity of describing your best friend is significantly
lower, the description complexity $K(e)$ drops while the causal complexity
$K_W (e)$ remains unchanged. This is why this latter  event appears unexpected.

Using these insights from AIT, we define the memorability $M(e)$ of an event as the absolute difference between the description complexity $K_d(e)$ of an event and its expected description complexity $K_{exp}(e)$:
\begin{equation}
  \label{eq:memorability}
  M(e) = |K_{exp}(e) - K_d(e)|
\end{equation}
Contrary to the definition of unexpectedness from Equation \ref{eq:unexpected}, we use an absolute value: we do so to acknowledge the fact that events more complex than expected can be memorable as well\footnote{In the original paper \cite{dessalles2011coincidences}, exceptionally complex events are described by considering complexity itself as a way to describe the event: see ``the Pisa Tower effect''\cite{dessalles_pisa_nodate}}. In the next section, we define computational approximations for the description complexity $K_d$ and the expected complexity $K_{exp}$ of events.

@@

%Computing the memorability of events begins with the introduction of formal definitions of events, how they are stored, how we can describe them and how these descriptions can be used to retrieve them.
We define \emph{events} as data points augmented with a \emph{label} indicating their nature (temperature event, failure event, addition/removal of a device) and a timestamp of occurrence. Formally:
\begin{equation}
  \label{eq:event}
  e = (l, t,\mathcal{D})
\end{equation}
where $l$ is the label, $t$ the timestamp and $\mathcal{D}$ a multi-dimensional data point representing the various characteristics of $e$: its duration, the maximum temperature reached, the sensor name, its position, etc. Labels can also be considered as classes of events, of which each event is a particular instance.

To model how humans are able to describe events by using qualifiers, we use \emph{predicates}: Boolean functions operating on events and, possibly, additional parameters: $\pi(e, a_1, a_2, \dots, a_n) \mapsto \{O,1\}$ is a predicate of arity $n$ operating on event $e$. In the rest of this paper, we will prefer the equivalent notation $\pi(e, k) \mapsto \{0,1\}$, where $k$ is a binary string encoding the sequence of arguments $a_1, \dots, a_n$. Using this notation, the predicate $\pi$ becomes a boolean function operating on $\mathbf{E} \times \{0,1\}^*$:
\begin{equation}
  \label{eq:predicate}
  \pi : \begin{cases}
\mathcal{M}\times \{0,1\}^{*} &\mapsto \{0,1\} \\
(e, k) &\mapsto \pi_{k}(e)
\end{cases}
\end{equation}
As an example of predicate, consider $\pi = \mathtt{year}$ and $k$ a string encoding the number $1$, thus
constructing the predicate $\mathtt{year(}e, 1\mathtt{)}$, which tells whether
the event $e$ occurred $1$ year ago.

As events occur, they are stored in the  \emph{base memory} $M_0$. As they are not directly comparable, the memory $M_0$ can be considered as having the structure of an unordered set. We denote by $\mathcal{M}$ the subsets of $M_0$. By extension, elements of $\mathcal{M}$, i.e. subsets of $M_0$, are also called \emph{memories}.

By applying a given predicate $\pi$ to all events contained in a memory $M \subseteq M_0$, and selecting only events satisfying $\pi$, one gets another memory $M_1 \subseteq M \subseteq M_0$. We call this operation a \emph{filter}:
\begin{equation}
  \label{eq:filter}
f_{\pi, k}: \begin{cases}
\mathcal{M} & \mapsto \mathcal{M}             \\
M           & \mapsto \{e \in M | \pi(e) \}
\end{cases}
\end{equation}
For instance, using the same $\pi = \mathtt{year}$ and $k=1$ as above, we can
build the filter $f_{\pi, k} = \mathtt{last\_{}year}$, which selects all events
that occurred last year.

As the output of a filter applied to a memory $M$ is another memory
object $M' \subseteq M$, we can compose filter
functions. A sequence of such filters is called a \emph{retrieval path}
\begin{equation}
\label{eq:ret_def}
p = (f_{\pi_{1}, k_{1}}, \dots, f_{\pi_{n}, k_{n}})
\end{equation}
By definition
$p(M) = f_{\pi_{n}, k_{n}}(\dots(f_{\pi_{1}, k_{1}}(M)))$.
In case the result of the operation $p(M)$ contains a single element
$e$, we say that the path $p$ \emph{retrieves} the element $e$ from $M$, and write
$p(M) = e$. In the example shown in Figure \ref{fig:filters}, the three filters $f_1, f_2, f_3$ form a retrieval path retrieving the event ``last year's hottest day'' from the base memory $M_0$.

@@

By considering only a finite number of possible predicates $\pi$ and arguments $k$, and a maximum path length, we can construct a finite set $P$ of possible retrieval paths. By limiting the search over this set, we get an upper bound of description complexity, and use this upper bound as an approximation:

@@

The approximation of description complexity from Equation \ref{eq:approx_k_desc} allows for
a direct implementation, which is shown in Algorithm \ref{alg:complex_iter}. This
algorithm operates iteratively: starting from the base memory $M_0$
(line 1), we apply all possible predicate concepts $\pi$ from a given finite set
$\Pi$ and programs $k$ (lines 6-7), up to a given length $\mathtt
{max\_len}$ bits, and apply them: $M' = f_{\pi, k}(M)$ (line 12). We then store
the pairs $(M', \mathtt{len(}\pi, k\mathtt{)})$ in an array $\mathtt{future_
{explore}}$. At the end of the iteration, the results of the filters become the
memories which will be explored during the next iteration(lines 21--23). Each
pass thus explore retrieval paths of increasing length. When a singleton memory
is reached, the complexity of its unique element is upper-bounded with the
length of the corresponding retrieval path (line 14).

\subsection{Computing Memorability}
As stated in Equation \ref{eq:memorability}, we define memorability $M(e)$ as the absolute difference between the description complexity of an event and its expected value. As we've just defined $K_d(e)$ and provided an approximation in Equation \ref{eq:approx_k_desc}, we now focus on defining the \emph{expected} description complexity of an event, $K_{exp}(e)$ that appears in Equation \ref{eq:memorability}

$K_{exp}(e)$ evaluates the complexity the user, or the system, would expect the occurrence of event $e$ to have, based on their previous knowledge. In our framework, this prior knowledge consists of the base memory $M_0$. The expected complexity of the event $e$ can be computed with a simple first-order approximation, i.e. estimating the average complexity of ``similar events'' over the base memory $M_0$.

Still, there is a difficulty in defining what
should be considered \emph{similar} events. Given that we deal with non comparable events, we may define the notion of similarity by referring once again to \emph{predicates}.For a given event $e$ and a given predicate $\pi_k$, we define a $\pi_k$-neighborhood of $e$ as the set $N_{\pi, k}(e)$ of all other events satisfying $\pi_k$.

@@

This definition is consistent with the intuitive idea that more similar events should weigh more in the computation. Indeed, if $e'$
is very similar to $e$, it will appear in many neighborhoods, since it
satisfies most of the predicates that $e$ satisfies. Therefore, it will
be present in more terms in Equation \ref{eq:expected}, and will weigh more in
the final result.

\subsection{From memorability to abduction}
% On peut étendre la définition utilisée pour la surprise en faisant intervenir
% les prédicats'' `sympathiques` par rapport à ce que l'on questionne.
Abductive inference builds on the computation
of the memorability score. \emph{Knowing} that we want to find a cause $c$
for an observed effect $e$, we try to find the most remarkable event in past
memory that is related to $e$. While our ``memorability'' score identifies remarkable past events, it does not take into account their relatedness to $e$.

The knowledge attached to the occurrence of $e$ can be integrated into the description complexity definition by using conditional complexity $K_d(c | e)$: The information contained in $e$ is considered as given, and therefore ``free'' in terms of complexity. For instance, when looking for a cause for
an anomaly in the living-room, other anomalies occurring in the same
living-room will be simpler, as the location ``living-room'' is already known
from the observation of the current anomaly.
